---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Prof. Dr. Ulrich Matter"
date: "06/12/2018"
output:
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
  pdf_document: default
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes_hsg.css
subtitle: 'Exercise 5: Applied Data Analysis with R'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, message=FALSE, warning=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```

# Exercises^[Adopted from https://r4ds.had.co.nz/tidy-data.html.]


## Exercise A
Load `tidyverse` and read `stocks.csv` into R (store it in variable `stocks`). The dataset is currently in long format. Use `spread()` to transform it into wide format (separate columns for different years). Store the resulting `tibble`/`data.frame` in a variable called `stocks2`. Then, run the following code.

```{r eval=FALSE}
gather(stocks2, "year", "return", `2015`:`2016`) == stocks
```

Why is the result not identical to the original dataset (`stocks`)? In other words, Why are `gather()` and `spread()` not perfectly symmetrical? (Hint: look at the variable types and think about column names.)

```{r echo=TRUE, purl= FALSE, message=FALSE, warning=FALSE}
# steps as instructed:
library(tidyverse)
stocks <- read_csv("../data/stocks.csv")
stocks_wide <- spread(stocks, year, return)
stocks_wide_long <- gather(stocks_wide, "year", "return", `2015`:`2016`) 

# test if symmetrical
stocks_wide_long == stocks
identical(stocks_wide_long, stocks)

# analyze the problem problem
str(stocks_wide_long)
str(stocks)

```



## Exercise B
Load `tidyverse` and read `countries.csv` into R (store it in variable `countries`). Then, run the following code.

```{r eval=FALSE}
gather(countries, 1999, 2000, key = "year", value = "cases")
```

Why does this code fail?

```{r eval=FALSE, purl= FALSE}

# sample solution
countries <- read_csv("../data/countries.csv")
# try the example code
gather(countries, 1999, 2000, key = "year", value = "cases")

# fix the problem (numeric values can't be used as variable names!)
gather(countries, `1999`, `2000`, key = "year", value = "cases")

```



## Exercise C
Load `tidyverse` and read `people.csv` into R (store it in variable `people`). Try to spread `people`. Why does the spreading fail? How could you add a new column to fix the problem?

```{r eval= FALSE, purl= FALSE}
# sample solution
people <- read_csv("../data/people.csv")
spread(people, key, value)

# look at the dataset
View(people)
str(people)

# key-variable (identifiers) contains duplicate values!

# how to fix this?
# assuming there are two Philip Woods
people$id <- c(1,1,2,3,3)
spread(people, key, value)

```


## Exercise D
Load `tidyverse` and read `preg.csv` into R (store it in variable `preg`). Tidy `preg`. Do you need to spread or gather it? What are the variables?

```{r echo=FALSE, purl= FALSE}
# sample solution
preg <- read_csv("../data/preg.csv")
preg # have a look at the data

# use common sense ('heuristics') to figure out what is what
# this could make sense:
gather(preg, "male", "female", key="gender", value = "count")
```




# References
