---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Prof. Dr. Ulrich Matter"
date: "05/12/2019"
output:
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
  pdf_document: default
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes_hsg.css
subtitle: 'Exercise 5: Applied Data Analysis with R'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, message=FALSE, warning=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```

# Exercises^[Adopted from https://r4ds.had.co.nz/tidy-data.html.]


## Exercise A
Load `tidyverse` and read `stocks.csv` into R (store it in variable `stocks`). The dataset is currently in long format. Use `spread()` to transform it into wide format (i.e., separate columns for different years). Store the resulting `tibble`/`data.frame` in a variable called `stocks2`. Then, run the following code.

```{r eval=FALSE}
gather(stocks2, "year", "return", `2015`:`2016`) == stocks
```

Why is the result not identical to the original dataset (`stocks`)? In other words, why are `gather()` and `spread()` not perfectly symmetrical? (Hint: look at the variable types and think about column names.)

```{r echo=TRUE, purl= FALSE, message=FALSE, warning=FALSE}
# steps as instructed
library(tidyverse)
stocks <- read_csv("../data/stocks.csv")
stocks2 <- spread(stocks, year, return)
stocks_wide_long <- gather(stocks2, "year", "return", `2015`:`2016`) 

# test if symmetrical
stocks_wide_long == stocks
# another way to test whether the datasets are exactly equal
identical(stocks_wide_long, stocks)
```

In order to evaluate the differences between the two datasets `stocks` and `stocks_wide_long`, take into account the variable types and the ordering of the columns.

```{r eval=FALSE}
# investigate the variable types
str(stocks_wide_long)
str(stocks)

```



## Exercise B
Load `tidyverse` and read `countries.csv` into R (store it in variable `countries`). Then, run the following code.

```{r eval=FALSE}
gather(countries, 1999, 2000, key = "year", value = "cases")
```

Why does this code fail?

```{r eval=FALSE, purl= FALSE}

# read the data
countries <- read_csv("../data/countries.csv")

# try the example code
gather(countries, 1999, 2000, key = "year", value = "cases")

```
The code fails because numeric values cannot be used as variable names.

```{r eval=FALSE, purl= FALSE}
# fix the problem 
gather(countries, `1999`, `2000`, key = "year", value = "cases")

```
A note on the order of a function's arguments: Compared to exercise A, the arguments of the function `gather` are shown in a different order. A deviation from the predefined order (as applied in exercise A) requires the addition of the specific names of the arguments (e.g., key = ...).


## Exercise C
Load `tidyverse` and read `people.csv` into R (store it in variable `people`). Try to spread `people`. Why does the spreading fail? How could you add a new column to fix the problem?

```{r eval= FALSE, purl= FALSE}
# read the data
people <- read_csv("../data/people.csv")
spread(people, key, value)

# look at the dataset
view(people)
str(people)
```
The issue is that the key-variable -- the identifier -- contains duplicate values.

```{r eval= FALSE, purl= FALSE}

# fix problem:
# assume there are two Phillip Woods and mark them with a different identifier
people$id <- c(1,1,2,3,3)
spread(people, key, value)

```


## Exercise D
Load `tidyverse` and read `preg.csv` into R (store it in variable `preg`). Tidy `preg`. Do you need to spread or gather it? What are the variables?

```{r eval=FALSE, purl= FALSE}
# read the data
preg <- read_csv("../data/preg.csv")

# look at the dataset
view(preg)

# use common sense ('heuristics') to figure out what information are contained
gather(preg, "male", "female", key = "gender", value = "count")
```




# References
