---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Exercise/Workshop 3:<br>Programming with Data'
author: "Prof. Dr. Ulrich Matter"
date: "28/10/2021"
output:   
  ioslides_presentation:
    css: ../../style/ioslides.css
    template: ../../style/nologo_template.html
    mathjax: "default"
logo: ../../img/logo.png
bibliography: ../../references/datahandling.bib
---




```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
library(knitr)
```

<style>
pre {
  font-size: 21px;
}
</style>

# Exercises

## Exercise A: Write a Sum Function


```{r eval=FALSE, purl=FALSE}
my_sum <- 
     function(x){
          
       return()
     }

```
Test your function with the vector `numbers = c(1, 2, 3)`.



## Exercise B: Robustness and Warnings
Once you have implemented and successfully tested `my_sum()`, see what happens if you call it with the following vector as input

```{r}
numbers2 <- c("1", "2", "3")
```

Why do we get an error? Investigate by comparing the vector `numbers` with the new vector `numbers2` (the functions `str()` and `class()` might help you).

## Exercise B: Robustness and Warnings

Extend your `my_sum()` function with a control statement that checks whether the input is of the right class for the function to work (Hint: `class(x)!="numeric"`). The goal is to make the function more robust to wrong input. That is, if the input is wrong, issue a warning but don't break down with an error. Use the following code for the warning message:

```{r eval=FALSE, purl=FALSE}
warning("Wrong input! Function only accepts numeric or integer values!")
```

Test your enhanced sum function with the two input vectors `numbers` and `numbers2`.




## Exercise C: Standard Deviation Function
During the lecture on Programming with Data we have step-by-step implemented our own R function to compute the mean. Now implement an R function called `my_sd` that computes the standard deviation, defined as: $SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$. (Hint: revisit the math operators section in the notes on the first exercise session: https://umatter.github.io/courses/datahandling/notes/html/01_tools_textfiles.html).

Test your implementation by comparing its output with the already implemented `sd()` function.



# Mock Exam Questions

# More Programming Exercises


## Exercises D and E

### Exercise D: Standard Error Function
Building on the function written in Exercise C, implement a function to compute (the estimate of) the sample mean standard error as defined here: $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$. Hint: the idea is to use your own implementation (in Exercise C) to compute $SD$ as a first step of your standard error function.

### Exercise E: t-test
Implement a function for the one-sample t-test.

$t = \frac{\bar{x} - \mu_0}{\frac{SD}{\sqrt{n}}}$.

See here: https://en.wikipedia.org/wiki/Student%27s_t-test.

## A Note on Breaking up Complex Problems

Exercises C and D should illustrate how you can approach more complex problems by breaking them up in smaller problems and addressing each of these step-by-step. That is, when looking at the problem of computing $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$, you see that computing $SD$ is a part of the overall problem, which can be addressed separately.
