---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: "Lecture 3: Data Processing Basics"
author: "Dr. Aurélien Sallin"
author-title: "Lecturer"
date: today
date-format: "MMMM YYYY"
params:
  show_code: true  # Toggle this to switch between student and solution versions
format:
  html:
    theme: ../../style/hsgtheme_reports.scss
    include-after-body: ../../style/report-footer.html
    embed-resources: true
    self-contained: true
    self-contained-math: true
highlight-style: dracula
---

```{r}
#| label: setup
#| include: false
options(width = 100)
knitr::opts_chunk$set(results = "hold")
library(knitr)
library(purrr)
```


# Handout Lecture 3

In this applied session, we will learn two important lessons: 

  1. All data and commands are stored as text files.
  2. Encoding matters.


## Encoding matters

- Store ["Hello World"]{.green-bold} in `helloworld.txt`.
  - Allocation of a block of computer memory containing "Hello World!".
  - Simply a sequence of `0`s and `1`s...
  - `.txt` indicates to the operating system which program to use when opening this file.
  
- Encoding and format tell the computer how to interpret the `0`s and `1`s.


You can store the file either using your favorite text editor, or using the command line from your terminal (Linux/Mac) or command prompt (Windows). The terminal can be also used from RStudio in Terminal (next to the console). Note that your path should be correct!

```{bash, eval=FALSE, echo = TRUE}
echo "Hello World!" > helloworld.txt

# Given your path
# echo "Hello World!" > C:/Users/aurel/OneDrive/Bureau/helloworld.txt
```


## Inspect a text file

Interpreting `0`s and `1`s as text... using `bash` (scripting language to interact with the operating system):

```{bash, eval=FALSE, echo = TRUE}
cat helloworld.txt; echo

# Given your path
# cat C:/Users/aurel/OneDrive/Bureau/helloworld.txt; echo
```

```{bash, echo=FALSE}
cat ../../data/helloworld.txt; echo
```

In `R`, use the `readLines()` function:

```{r, eval=TRUE, echo = TRUE}
readLines("helloworld.txt")
```

Directly looking at the `0`s and `1`s... (this works only from the terminal, not from the R console):

```{bash, eval = FALSE}
xxd -b helloworld.txt

# Given your path, in the terminal
# xxd -b C:/Users/aurel/OneDrive/Bureau/helloworld.txt
```

```{bash, echo = FALSE, eval = FALSE}
xxd -b ../../data/helloworld.txt
```

```{}
00000000: 01001000 01100101 01101100 01101100 01101111 00100000  Hello 
00000006: 01010111 01101111 01110010 01101100 01100100 00100001  World!
```

  - `xxd` is a hex dump tool. It displays the raw bytes of a file in a readable format.
  - The option `-b` tells it to show those bytes in binary instead of the default hexadecimal.


```{r, eval=FALSE, echo = F}
# Inspecting a text file as binary in R
file_content <- readBin("../../data/helloworld.txt", 
                        what = "raw",
                        n = 12)

binary_content <- paste(as.character(
  sapply(file_content, function(x) {
    paste0(as.integer(rawToBits(x)), collapse = "")
  })), 
  collapse = " ")

binary_content
```


Similarly we can display the content in hexadecimal values:


```{bash, eval = FALSE, echo = TRUE}
xxd  data/helloworld.txt
```

```{bash, echo = FALSE, eval = FALSE}
xxd  ../../data/helloworld.txt
```

```{}
00000000: 4865 6c6c 6f20 576f 726c 6421            Hello World!
```

For instance, we see that `6c6c` is two times the 108 in decimal. This is Latin Small Letter L in unicode. It is the code point `U+006C` in unicode. In utf-8, this is translated as a single byte `0x6C`, or in binary `01101100`.

<br>


## Encoding issues

Import the file `hastamanana.txt` from Canvas. Read it with your text editor. Then check it with `cat` from the terminal or command prompt.


### What is the problem?

```{bash, eval=FALSE, echo = TRUE}
cat hastamanana.txt; echo
```

```{bash, eval=FALSE}
cat ../../data/hastamanana.txt; echo
```

```{}
## Hasta Ma?ana!
```


Using `xxd` to inspect the file in hex, we see that the problem can likely be found in the hex `f1`, which is the ñ. `f1` is the problematic byte. [It corresponds to the "ñ" in ISO-8859.](https://cs.stanford.edu/people/miles/iso8859.html) and is the code `U+00F1` in unicode (Latin-1 supplement). In UTF-8, f1 is not a valid character by itself, [as UTF-8 expects two bytes (c3 b1)](https://www.utf8-chartable.de/).

### Inspect the encoding

```{bash, eval = FALSE, echo = TRUE}
file -b hastamanana.txt
```

```{bash, echo = FALSE, eval = T}
file -b ../../data/hastamanana.txt
```
It gives "ISO-8859 text, with CRLF line terminators". This means that the file is encoded in ISO-8859-1 (also known as Latin-1), which is a character encoding standard. 


### Read the file again, this time with the correct encoding

```{bash, eval = FALSE, echo = TRUE}
iconv -f iso-8859-1 -t utf-8 hastamanana.txt | cat
```

```{bash, echo = FALSE}
iconv -f iso-8859-1 -t utf-8 ../../data/hastamanana.txt | cat
```




<br>


## All we see on the Internet is, ultimately, text files.

When you access a website, you are downloading a text file (the HTML source code of the webpage). When you upload a picture on Instagram, you are uploading a text file (the binary encoding of the image). When you send an email, you are sending a text file (the email content encoded in a specific format).

In this short illustration, we will see how to import an html content from the web. This will serve as an example to get an idea about a computing environment and the interaction between man and computer.

<br>

### Roadmap

  1. Access a website (over the Internet), use keyboard to enter data into a website. 

  2. R program accesses the data of the Google sheet (again over the Internet), downloads the data, and loads it into RAM.

  3. Data processing: produce output (in the form of statistics/plots), output on screen.
  
  
<br>

## Load necessary packages

We will need the packages `httr` and `pryr`. `httr` is used for working with URLs and HTTP. `pryr` is used to inspect the size of objects in RAM.

```{r warning=FALSE}
library(httr)
library(pryr)
```

We want to load the website "https://www.economist.com/". We use the `GET()` function from the package `httr` to save the website in our environment. We can access the structure of the object with the function `str()` for "structure".

```{r}
economist <- GET("https://www.economist.com/")
```

We see that we have the following data: (1) the url written by hand in the form of a string, (2) the source code HTML being sent through the network, and (3) the HTML document loaded in the RAM. On top of this, we can see other meta-data, such as cookies, headers, date of access, etc.

We can ask the size of the object in the RAM using `object_size()` from package `pryr`.

```{r}
pryr::object_size(economist)
```


<br>

The html text data was sent in hexadecimal byte values to our computer. This is visible using the command `content` on our query, specifying that we want to see the raw data.

```{r}
economistRaw <- content(economist, as = "raw")
head(economistRaw, 15)
```

When translated to ASCII characters, it represents a string typically found at the beginning of HTML documents.

Hexadecimal  | Character 
-----|-----
3c | <
21 | !
44 | D
4f | O
43 | C
54 | T
59 | Y
50 | P
45 | E
20 | (a space character)
68 | h
74 | t
6d | m
6c | l

<br>

Now we can encode these hexadecimal values using the command `content(., as = "text")`.

```{r, collapse=TRUE}
economistText <- content(economist, as = "text")
head(economistText)
```
<br>

Finally, we can tell R to store the source code of the downloaded webpage into our local hard drive. 

```{r, collapse=TRUE}
writeLines(economistText, "economistText.html")
```