---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Ulrich Matter"
date: "14/10/2021"
output:
  html_document:
    highlight: tango
    mathjax: //mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML
    theme: cerulean
    css: ../../style/notes_hsg.css
  pdf_document: default
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
subtitle: 'Exercise/Workshop 2: Computer Code and Data Storage'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, message=FALSE, warning=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```

# Repetition: Data Structures in R

R provides different object classes to work with data. In simple terms, they differ with regard to how data is structured when saved in a variable. This, in turn, defines how we can access different parts of the data in this variable. For example, if we store a bunch of numbers in a numeric vector, we can then access each of these numbers individually, by telling R to give us the first, second, nth element of this vector. If we store data in a two-dimensional structure such as a matrix, we can then access the data column-wise, row-wise, or cell-wise. Depending on the task, one or the other object class is preferable to store and work with data.

## Vectors and Lists
```{r}
# A vector containing numeric (or integer) values
numeric_vector <- 10:20
numeric_vector[2]
numeric_vector[2:5]

# A string vector ('a vector containing text')
string_vector <- c("a", "b", "c")
string_vector[-3]

# Lists
# A list can contain different types of elements, for example a numeric vector and a string_vector
mylist <- list(numbers = numeric_vector, letters = string_vector)
mylist

# We can access the elements of a list in various ways
# with the element's name
mylist$numbers
mylist["numbers"]
# via the index
mylist[1]
# with [[]] we can access directly the content of the element
mylist[[1]]

# lists can also be nested (list of lists of lists....)
mynestedlist <- list(a = mylist, b = 1:5)
```

## Matrices and Data Frames
```{r}

# matrices
mymatrix <- matrix(numeric_vector, nrow = 4)
# get the second row
mymatrix[2,]
# get the first two columns
mymatrix[, 1:2]

# data frames ("lists as columns")
mydf <- data.frame(Name = c("Alice", "Betty", "Claire"), Age = c(20, 30, 45))
mydf
# select the age column
mydf$Age
mydf[, "Age"]
mydf[, 2]
# select the second row
mydf[2,]


```

## Classes and Data Structure
When importing data to R, using a new R-package, or extracting data from web sources, it is imperative to understand how the data is structured in the R-object (variable) one is working with. A lot of errors occur due to giving R functions input in the wrong format/structure. Having a close look at how the data stored in a variable that we want to work with further can help avoid such errors.^[Note that basic R functions as well as functions distributed via R packages are usually very well documented (look up documentation with `help(FUNCTION-NAME)` or `?FUNCTION-NAME`. An important part of such documentations is what class the input arguments of a function must have and what the class of the returned object is. See, for example the documentation of `mean()` by typing `?mean` into the R-console and hit enter.] The function `class()` returns the class(es) of an R object:

```{r}
# have a look at what kind of object you are dealing with
class(mydf)
class(mymatrix)

```

To get a clear idea of the structure of data stored in a specific variable, we can call the function `str()` (for structure) with the object of interest as the only function argument. This is particularly helpful when dealing with a complex/nested data structure. 

```{r}
# have a closer look at the data structure
str(mydf)
```


