---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Aurélien Sallin (lecturer) and Federica Mascolo (class tutor)"
date: "10/10/2024"
output:
  pdf_document: default
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes_hsg.css
subtitle: 'Exercise/Workshop 2: Function, iterations, a data structures'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, warning=FALSE, message=FALSE}
options(width = 100)
library(knitr)
```

# Exercises

## Exercise A: Write a Sum Function

In the lecture on Programming with Data, we have discussed the most basic programming concepts in R. Among other things, you have learned how we can use for-loops to sum up numbers stored in a numeric vector.

1. Use the code example in the respective lecture notes and the following empty function construct to implement a function that takes a numeric vector as input and returns the sum of the vector's elements.

```{r eval=FALSE, purl=FALSE}
my_sum <- function(x){
          
     }
```

2. Define the vector `numbers = c(1, 2, 3)`. Then, compare your function `my_sum()` with the `sum()` function (the pre-implemented sum function in R) when using `numbers` as the input argument: Do you get the same output for `my_sum(numbers)` and `sum(numbers)`?

## Exercise B: Robustness and Warnings

1. Once you have implemented and successfully tested `my_sum()`, call your function with the following vector as input. What happens?

```{r}
numbers2 <- c("1", "2", "3")
```

2. Why do we get an error? Investigate by comparing the vector `numbers` from Exercise A with the new vector `numbers2`. Hint: the functions `str()` and `class()` might help you.

3. Extend your `my_sum()` function with a control statement that checks whether the input is of the right class for the function to work (Hint: `class(x)!="numeric"`). The goal is to make the function more robust to wrong input. That is, if the input is wrong, issue a warning but don't break down with an error. Use the following code for the warning message:

```{r eval=FALSE, purl=FALSE}
warning("Wrong input! This function only accepts numeric or integer values!")
```

4. Test your enhanced sum function with the two input vectors `numbers` and `numbers2`.


## Exercise C: use the sum function and lapply/loop/map on a list of vectors
In this exercise, you will learn how to apply a custom function to multiple elements in a list using different methods.

1. use the my_sum() function created earlier 

2. create a list of numeric vectors
```{r eval=FALSE}
# Create a list of vectors
list_of_numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(7, 8, 9, 10)
)
```

3. Use the lapply() function to apply my_sum() to each vector in the list_of_numbers list. Assign the result to a new variable called sums_lapply. Print the result.

4. Use a for loop to achieve the same result as in Step 3. Initialize an empty vector called sums_loop and then iterate over each vector in list_of_numbers, storing the sums in sums_loop. Print the result.

5. Use purrr::map() to Apply my_sum() to Each Vector
Use the map() function from the purrr package to apply my_sum() to each vector in list_of_numbers. Assign the result to a variable called sums_map. Make sure to load the purrr package before using map(). Print the result.

## Exercise D: Data stuctures

### Vectors, factors and lists
1. Create a vector student_age containing the ages of 5 students.
   -  print its length and type using length() and typeof() functions

2. Create a factor student_grades representing the grades of the same 5 students: "A", "B", "A", "C", "B".
   - print the factor levels, why do you use factors?
  
3. Create a list student_info that combines student_ages, student_grades, and another vector student_names containing the names of the students.
   - access individual elements using both numeric and named indexing.

4. What is the difference between a vector and a factor? Why do you use a list instead of a vector when dealing with different types of data?

### Matrices and arrays
1. Create a matrix scores_matrix representing the scores of 3 students in 3 subjects
   - give row and column names using rownames() and colnames().
   - print the matrix and the score of a student that you prefer.
  
2. Create an array student_array to store the same data as in scores_matrix, but this time, add scores from another semester. Use the same students and subjects but with different scores.
   - this 3-dimensional array, access "Student 3" scores in the first subject listed across both semesters.

3. What are the differences between a matrix and an array? When would you use an array instead of a matrix?
  
### Data Frames, tibble and data.table

1. Create a data frame students_df combining all the data for 5 students: student_names, student_ages, student_grades, and another vector attendance containing their class attendance (in percentage): 85, 92, 78, 88, 90.
   - access the columns of attendance.
   - access the data for student 3.
  
2. Why is a data frame more suitable for this kind of structured data than a matrix?

3. Convert the existing students_df data frame into a tibble using the as_tibble() function.
   - name the tibble students_tibble and print the data frame and the tibble.


# Additional Exercises for Training 
5. Use the output from the previous exercise. Compare the syntax of common data manipulation tasks (e.g., filtering, grouping, summarizing) between data.frame, tibble, and data.table.
   - Filter students with attendance greater than 85
   - Calculate mean age of students who passed


## Exercise E: Standard Deviation Function

1. During the lecture on Programming with Data we have step-by-step implemented our own R function to compute the mean. Now implement an R function called `my_sd` that computes the standard deviation, defined as: $SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$. Hint: revisit the math operators section in the notes on the first exercise session.

2. Test your implementation by comparing its output with the already implemented `sd()` function.


## Exercise F: Standard Error Function

Building on the function written in Exercise D, implement a function to compute (the estimate of) the sample mean standard error as defined here: $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$. Hint: the idea is to use your own implementation (in Exercise D) to compute $SD$ as a first step of your standard error function.

Exercises D and E should illustrate how you can approach more complex problems by breaking them up in smaller problems and addressing each of these step-by-step. That is, when looking at the problem of computing $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$, you see that computing $SD$ is a part of the overall problem, which can be addressed separately.

## Exercise G: T-test

Implement a function for the one-sample [t-test](https://en.wikipedia.org/wiki/Student%27s_t-test):
$t = \frac{\bar{x} - \mu_0}{SE}$.