---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Ulrich Matter"
date: "22/10/2020"
output:
  pdf_document: default
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes.css
subtitle: 'Exercise/Workshop 3: Programming with Data'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, warning=FALSE, message=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```


# Exercises

## Exercise A: Write a Sum Function

In the lecture on Programming with Data, we have discussed the most basic programming concepts in R. Among other things, you have learned how we can use for-loops to sum up numbers stored in a numeric vector.

1. Use the code example in the respective lecture notes and the following empty function construct to implement a function that takes a numeric vector as input and returns the sum of the vector's elements.

```{r eval=FALSE, purl=FALSE}
my_sum <- 
     function(x){
          
     }
```

2. Define the vector `numbers = c(1, 2, 3)`. Then, compare your function `my_sum()` with the `sum()` function (the pre-implemented sum function in R) when using `numbers` as the input argument: Do you get the same output for `my_sum(numbers)` and `sum(numbers)`?

## Exercise B: Robustness and Warnings

1. Once you have implemented and successfully tested `my_sum()`, call your function with the following vector as input. What happens?

```{r}
numbers2 <- c("1", "2", "3")
```

2. Why do we get an error? Investigate by comparing the vector `numbers` from Exercise A with the new vector `numbers2`. Hint: the functions `str()` and `class()` might help you.

3. Extend your `my_sum()` function with a control statement that checks whether the input is of the right class for the function to work (Hint: `class(x)!="numeric"`). The goal is to make the function more robust to wrong input. That is, if the input is wrong, issue a warning but don't break down with an error. Use the following code for the warning message:

```{r eval=FALSE, purl=FALSE}
warning("Wrong input! This function only accepts numeric or integer values!")
```

4. Test your enhanced sum function with the two input vectors `numbers` and `numbers2`.


## Exercise C: Standard Deviation Function

1. During the lecture on Programming with Data we have step-by-step implemented our own R function to compute the mean. Now implement an R function called `my_sd` that computes the standard deviation, defined as: $SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$. Hint: revisit the math operators section in the notes on the first exercise session.

2. Test your implementation by comparing its output with the already implemented `sd()` function.

# Additional Exercises (not exam-relavant)

## Exercise D: Standard Error Function

Building on the function written in Exercise C, implement a function to compute (the estimate of) the sample mean standard error as defined here: $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$. Hint: the idea is to use your own implementation (in Exercise C) to compute $SD$ as a first step of your standard error function.

Exercises C and D should illustrate how you can approach more complex problems by breaking them up in smaller problems and addressing each of these step-by-step. That is, when looking at the problem of computing $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$, you see that computing $SD$ is a part of the overall problem, which can be addressed separately.

## Exercise E: T-test

Implement a function for the one-sample [t-test](https://en.wikipedia.org/wiki/Student%27s_t-test):
$t = \frac{\bar{x} - \mu_0}{\frac{SE}{\sqrt{n}}}$.


# Mock Exam Questions


```{r echo=FALSE, results='asis', message=FALSE}
library(examr)
library(pander)

# load file with questions
questions_file <- "../../../datahandling-examination/data/questions_db.xlsx"
my_questions <- read_questions(questions_file, year = 2018, mock = TRUE, topic = "week3-4")

# select questions
selection_mock_ex2 <- select_questions(my_questions)
questions_text <- generate_question_text(selection_mock_ex2)  
pander(questions_text, sep="\n")
```