---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: "Lecture 2: Students' notebook"
author: "Dr. Aurélien Sallin"
params:
  show_code: true  # Toggle this to switch between student and solution versions
format:
  html:
    theme: ../../style/hsgtheme.scss
    embed-resources: true
    self-contained: true
    self-contained-math: true
highlight-style: dracula
---

```{r}
#| label: setup
#| include: false
options(width = 100)
library(knitr)
library(purrr)
```


# Basic Programming Concepts in R

Elementary commands consist of either expressions or assignments. If an expression is given as a command, it is evaluated, printed (unless specifically made invisible), and the value is lost. An assignment also evaluates an expression and passes the value to a variable but the result is not automatically printed.

Commands are separated either by a semi-colon (‘;’), or by a newline. Elementary commands can be grouped together into one compound expression by braces (‘{’ and ‘}’). Comments can be put almost3 anywhere, starting with a hash mark (‘#’), everything to the end of the line is a comment.

https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Simple-manipulations-numbers-and-vectors

## Values and variables

  - Assign a variable in the environment using `<-`
  - Manipulate the variable

```{r}
2 + 2  # CTRL + ENTER

my_variable <- 10
class(my_variable)
objects()
rm(my_variable)

my_integer <- 9L
class(my_integer)

my_character <- "aurelien"
class(my_character)

my_logical <- TRUE
class(my_logical)
```

## Vectors

```{r}
vector_numeric <- c(1, 2, 3)
vector_numeric * 5
vector_numeric - 10
length(vector_numeric)
vector_numeric[1]
vector_numeric[3]
vector_numeric[c(1, 2)]
```

## Matrices

```{r}
v1 <- c(1, 2, 3, 4)
v2 <- c(10, 9, 8, 7)

m1 <- cbind(v1, v2)
print(m1)
#m1

m2 <- rbind(v1, v2)
m2

v3 <- c(2, 4, 6)
v4 <- c(10, 7)
# Note: differing lengths will recycle; try cbind(v3, v4) and read the warning
# cbind(v3, v4)
```

### Comparisons

```{r}
store_value1 <- 1
store_value2 <- 7
store_value1 == store_value2
```

# Operators

## Math operators: basic arithmetic

```{r}
2 + 2
4 * 5
20 / 5
sum_result <- 2 + 2
sum_result - 2

# Modulo (remainder)
5 %% 3

# Integer division
5 %/% 3
```

## Assignment and comparison operators

```{r}
# Assignment
b <- 1 + 1
3 -> my_var

# Avoid global assignment in this course
# b <<- 1 + 1

# Comparison
1 == 2
1 != 2
1 > 2
1 < 2
1 >= 2
1 <= 2
```

## Logical and miscellaneous operators

```{r}
# Logical
TRUE & FALSE
TRUE | FALSE
!TRUE

# Miscellaneous
1:10
2 %in% c(1, 2)
```

## Other common math functions

```{r}
4^2
sqrt(4^2)
log(2)
exp(10)
log(exp(10))
```

# Loops

## for-loop

```{r}
# Number of iterations
n <- 20
for (i in c(2, 4, 6, 8)) {
  print(i)
}

vector_loop <- c("brian", "mark", "sophia")
for (i in vector_loop) {
  sentence <- paste(i, "likes icecream.")
  print(sentence)
}
```

## while-loop

```{r}
x <- 1
while (x <= 10) {
  print("88")
  x <- x + 1
}
```

# Logical and control statements

```{r}
2 + 2 == 4
3 + 3 == 7
4 != 7
6 > 3
6 < 7
6 <= 6
```

```{r}
condition <- TRUE
if (condition) {
  print("This is true!")
} else {
  print("This is false!")
}
```

# Functions

## What is a function?

A function maps inputs to outputs. In R, functions can be built-in (from base or packages) or user-defined.

```{r}
# install.packages("<PACKAGE NAME>")
# library(<PACKAGE NAME>)
```

## Three components of a function

```{r}
myfun <- function(x, y) {
  z <- x + y
  return(z)
}
```

```{r}
formals(myfun)
```

```{r}
body(myfun)
```

```{r}
environment(myfun)
```

## Example: a simple power function

```{r}
powerFunction <- function(base, exponent) {
  results <- base ^ exponent
  return(results)
}

powerFunction(exponent = 2, base = 3)
powerFunction(base = 2, exponent = 3)
powerFunction(2, 3)
powerFunction(c(2, 4, 3), 3)
```

# Step up your game: functionals

A functional is a function that takes a function as input and returns a vector or list.

## Using lapply and purrr::map

```{r}
# User-defined function
triple <- function(x) {
  y <- x * 3
  return(y)
}

triple(1); triple(2); triple(3); triple(4)

# Base R
t1 <- lapply(1:4, triple)
t1

# purrr
map_dbl(1:4, triple)
map(1:4, triple)
```

## apply over matrices

```{r}
# Empty matrix with 2 rows and 4 columns
mymatrix <- matrix(c(1, 2, 3, 11, 12, 13, 1, 10), nrow = 2, ncol = 4)
mymatrix

# Row sums with a for-loop
for (i in 1:2) {
  s <- sum(mymatrix[i, ])
  print(s)
}

# Row sums with apply
apply(mymatrix, MARGIN = 1, sum)
```

# Tutorials

## Tutorial 1: a function to compute the mean

Starting point: the mean is

$\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i.$

Write a function `meaN(x)` that computes the mean of a numeric vector.

```{r}
meaN <- function(x) {
  # your code
}
```

\::: {.callout-note collapse="true" title="Solution"}

```{r}
meaN <- function(x){
  m <- sum(x) / length(x)
  return(m)
}

meaN(c(1, 2, 3))
vector <- c(1, 3, 5, 3, 2)
meaN(vector)
```

\:::

## Tutorial 2: on slow and fast sloths

We simulate two sloth populations over time. At the beginning of year 1 there are 1000 slow sloths and 1 fast sloth. Each year, each sloth has one offspring. Slow sloths have 40% mortality; fast sloths 30% mortality. Find the first year when fast sloths outnumber slow sloths.

```{r}
# your code
```

\::: {.callout-note collapse="true" title="Solution"}

```{r}
i <- 1
Nslow <- 1000
Nfast <- 1

while (Nslow > Nfast) {
  Nslow <- (Nslow * 2) * 0.6
  Nfast <- (Nfast * 2) * 0.7
  i <- i + 1
}

i

# Optional check
Nslow <- 1000; Nfast <- 1
for (k in 1:48) {
  Nslow <- c(Nslow, (Nslow[k] * 2) * 0.6)
  Nfast <- c(Nfast, (Nfast[k] * 2) * 0.7)
}
# tail(Nslow); tail(Nfast)
```

\:::

## Tutorial 3: a loop function

Create a function that repeatedly appends the sum of the current last three elements of a vector `lst` to `lst`, looping 25 times. Hint: use `append()` and `tail()`.

```{r}
appendsums <- function(lst) {
  # your code
}

# Example to test:
# sum_three <- c(0, 1, 2)
# appendsums(sum_three)
# After running, check: sum_three[10] == 125
```

\::: {.callout-note collapse="true" title="Solution"}

```{r}
appendsums <- function(lst){
  for (i in 1:25) {
    lst <- append(lst, sum(tail(lst, 3)))
  }
  lst
}

sum_three <- c(0, 1, 2)
sum_three <- appendsums(sum_three)
sum_three[10]
```

\:::

# Appendix: quick reference

* Vector constructors: `c()`, `seq()`, `rep()`
* Matrix builders: `matrix()`, `cbind()`, `rbind()`
* Apply family: `lapply()`, `sapply()`, `vapply()`, `apply()`
* purrr mappers: `map()`, `map_lgl()`, `map_int()`, `map_dbl()`, `map_chr()`
* Subsetting: `x[i]`, `x[i, j]` for matrices, logical masks
* Comparison: `==`, `!=`, `<`, `<=`, `>`, `>=`
* Logical: `&`, `|`, `!`
