---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Prof. Dr. Ulrich Matter"
date: "05/12/2019"
output:
  pdf_document: default
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes_hsg.css
subtitle: 'Exercise 5: Applied Data Analysis with R'
bibliography: ../../references/datahandling.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, warning=FALSE, message=FALSE}
options(width = 100)
library(knitr)
library(bookdown)
```



# Exercises^[Adopted from https://r4ds.had.co.nz/tidy-data.html.]

Download and unzip `data.zip` (posted along these instructions on StudyNet). The zip-file contains four csv-files with data needed to solve the following exercises. Make sure you store them in the respective working directory of your R session when solving the exercises.

```{r echo=FALSE, message=FALSE, purl=FALSE, warning=FALSE}
# generate csvs for exercises (all taken from https://r4ds.had.co.nz/tidy-data.html)
library(tidyverse)
# 1.
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
write_csv(stocks, "../data/stocks.csv")

# 2. 
write_csv(table4a, "../data/countries.csv")

# 3. 
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
write_csv(people, "../data/people.csv")

# 4. 
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
write_csv(preg, "../data/preg.csv")

```


## Exercise A
Load `tidyverse` and read `stocks.csv` into R (store it in variable `stocks`). The dataset is currently in long format. Use `spread()` to transform it into wide format (separate columns for different years). Store the resulting `tibble`/`data.frame` in a variable called `stocks2`. Then, run the following code.

```{r eval=FALSE}
gather(stocks2, "year", "return", `2015`:`2016`) == stocks
```

Why is the result not identical to the original dataset (`stocks`)? In other words, Why are `gather()` and `spread()` not perfectly symmetrical? (Hint: look at the variable types and think about column names.)

```{r echo=FALSE,  purl= FALSE, message=FALSE}
stocks <- read_csv("../data/stocks.csv")
stocks2 <- spread(stocks, year, return)
#gather(stocks2, "year", "return", `2015`:`2016`) == stocks
```



## Exercise B
Load `tidyverse` and read `countries.csv` into R (store it in variable `countries`). Then, run the following code.

```{r eval=FALSE}
gather(countries, 1999, 2000, key = "year", value = "cases")
```

Why does this code fail?

```{r echo=FALSE, eval=FALSE, purl= FALSE}
# sample solution
countries <- read_csv("../data/countries.csv")
#gather(countries, `1999`, `2000`, key = "year", value = "cases")

```



## Exercise C
Load `tidyverse` and read `people.csv` into R (store it in variable `people`). Try to spread `people`. Why does the spreading fail? How could you add a new column to fix the problem?

```{r eval=FALSE, purl= FALSE}
# sample solution
people <- read_csv("people.csv")
# assuming there are two Philip Woods
people$id <- c(1,1,2,3,3)
spread(people, key, value)

```


## Exercise D
Load `tidyverse` and read `preg.csv` into R (store it in variable `preg`). Tidy `preg`. Do you need to spread or gather it? What are the variables?

```{r eval=FALSE, echo=FALSE, purl= FALSE}
# sample solution
preg <- read_csv("preg.csv")
gather(preg, "male", "female", key="gender", value = "count")
```



# Mock Exam Questions



```{r echo=FALSE, results='markdown', message=FALSE, warning=FALSE}

library(pander)
library(examr)
# load file with questions
questions_file <- "~/datahandling-examination/data/questions_db.xlsx"
my_questions <- read_questions(questions_file, year = 2018, mock = TRUE, topic = "week7-8")

# select questions
selection_mock_ex2 <- select_questions(my_questions)
questions_text <- generate_question_text(selection_mock_ex2)  
pander(questions_text, sep="\n")

```




