---
title: 'Data Handling: Import, Cleaning and Visualisation'
author: "Aurélien Sallin (lecturer) and Federica Mascolo (class tutor)"
date: "10/10/2024"
output:
  pdf_document: default
  html_document:
    highlight: tango
    mathjax: http://example.com/mathjax/MathJax.js?config=TeX-AMS-MML_HTMLorMML
    theme: cerulean
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{hyperref}
css: ../../style/notes_hsg.css
subtitle: 'Example Solution for Exercise/Workshop 2: Computer Code and Data Storage'
bibliography: ../../references/datahandling.bib
---

```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE, warning=FALSE, message=FALSE}
options(width = 100)
library(knitr)
library(bookdown) 


```

# Exercise solutions

## Exercise A: Write a Sum Function

In the lecture on Programming with Data, we have discussed the most basic programming concepts in R. Among other things, you have learned how we can use for-loops to sum up numbers stored in a numeric vector.

1. Use the code example in the respective lecture notes and the following empty function construct to implement a function that takes a numeric vector as input and returns the sum of the vector's elements.

```{r eval=TRUE, purl=FALSE}
my_sum <- function(x){
  
  # Initialize the total - we start at 0
  total <- 0
  
  for (num in x){
      total <- total + num
    }
  
  return(total)
}

```

2. Define the vector `numbers = c(1, 2, 3)`. Then, compare your function `my_sum()` with the `sum()` function (the pre-implemented sum function in R) when using `numbers` as the input argument: Do you get the same output for `my_sum(numbers)` and `sum(numbers)`?


*We get the same result in both cases - our function seems to be working:*

```{r}
numbers = c(1,2,3)

my_sum(numbers)

sum(numbers)
```
## Exercise B: Robustness and Warnings

1. Once you have implemented and successfully tested `my_sum()`, call your function with the following vector as input. What happens?

```{r}
numbers2 <- c("1", "2", "3")
```

*In this case We get an error, R will not compute the sum and complains about a non-numeric argument:* `Error in total + num : non-numeric argument to binary operator'

```{r, eval=FALSE}
my_sum(numbers2)
```


2. Why do we get an error? Investigate by comparing the vector `numbers` from Exercise A with the new vector `numbers2`. Hint: the functions `str()` and `class()` might help you.


*We can see that the numbers2 vector we provided does not contain numeric values but strings, i.e. a series of characters:*
```{r}
# Investigate the structures
str(numbers)
str(numbers2)

# Check the classes
class_numbers <- class(numbers)
class_numbers2 <- class(numbers2)

print(paste("Class of 'numbers':", class_numbers))
print(paste("Class of 'numbers2':", class_numbers2))
```


3. Extend your `my_sum()` function with a control statement that checks whether the input is of the right class for the function to work (Hint: `class(x)!="numeric"`). The goal is to make the function more robust to wrong input. That is, if the input is wrong, issue a warning but don't break down with an error. Use the following code for the warning message:

```{r eval=FALSE, purl=FALSE}
warning("Wrong input! This function only accepts numeric or integer values!")
```


*We can fix our code by adding an if-statement to make sure we get a warning in case the input is not numeric. The `is.numeric()` command will return `TRUE` if x is numeric (and also if x is an integer, hence it is more flexible than requiring `class == "numeric"` for example)*
```{r}
my_sum <- function(x){
  total <- 0
  
  # Check for the right input class
  if (is.numeric(x)) {
    for (num in x){
      total <- total + num
    }
  } else {
    warning("Wrong input! This function only accepts numeric or integer values!")
    total <- NA_real_ # if no sum can be computed, we return a missing value 
  }
  
  return(total)
}
```


*Note that we would not want to use our own function to sum stuff up, because on top of not being able to address missing entries (easy to fix, see first exercise), it is a lot slower:*

```{r}
# Creating vector with 100 million random (normally distributed) numbers
big_vector = rnorm(100000000)
# Time the built-in sum function
time_builtin_sum <- system.time({
  sum(big_vector)
})[3]  # Extracting the elapsed time

# Time the custom my_sum function
time_my_sum <- system.time({
  my_sum(big_vector)
})[3]  # Extracting the elapsed time

# Print the results
print(paste("Execution time of built-in sum:", time_builtin_sum, "seconds"))
print(paste("Execution time of custom my_sum:", time_my_sum, "seconds"))

```



4. Test your enhanced sum function with the two input vectors `numbers` and `numbers2`.

*We can see that we get a warning now when running the function on the second vector. Importantly, the function also returns a missing value for the sum, such that we are aware that no valid result is available. Specifically, the function should not return the value of 0 we initially set for the `total` in the function above, because that may cause a lot of issues if interpreted as a proper result.*
```{r}
my_sum(numbers)
my_sum(numbers2)
```

## Exercise C: use the sum function and lapply/loop/map on a list of vectors

1. **`lapply()`**: Part of base R's "apply" family of functions, which are used for functional programming.
2. **`for` loops**: A more traditional, imperative style of programming where you explicitly loop over elements.
3. **`purrr::map()`**: A function from the `purrr` package that provides a more modern and consistent interface for functional programming in R, part of the tidyverse collection of packages.

```{r}

my_sum <- function(x){
  
  # Initialize the total - we start at 0
  total <- 0
  
  for (num in x){
    total <- total + num
  }
  
  return(total)
}

list_of_numbers <- list(
  c(1, 2, 3),
  c(4, 5, 6),
  c(7, 8, 9, 10)
)


```


```{r}
# lapply
sums_lapply <- lapply(list_of_numbers, my_sum)
print(sums_lapply)

# loop
sums_loop <- vector("list", length(list_of_numbers))  # Initialize an empty list
for (i in seq_along(list_of_numbers)) {
  sums_loop[[i]] <- my_sum(list_of_numbers[[i]])
}
print(sums_loop)

# map
library(purrr)
sums_map <- map(list_of_numbers, my_sum)
print(sums_map)

```

  - lapply() takes a list (or vector) and applies a function to each element, returning a list.You can use it when you want to apply a function to each element of a list and get a list back. Is more concise and faster than loops.
  
  - for loops explicitly iterate over each element of a list or vector, applying a function or operation. You can use it when you want to have more control of the operations. The logic is the same in other programming languages.
  
  - `map()` is part of the `purrr` package (part of the tidyverse), which provides a consistent set of functions. It applies a function to each element of a list and returns a list. Variants like `map_dbl()`, `map_int()`, etc., return vectors of specific types. You can us it if you are working with tidyverse or you want to return specific types.

## Exercise D: Data structures 

#### Vectors, Factors, and Lists

1. **Create a vector** `student_ages` containing the ages of 5 students: `18, 21, 20, 22, 19`.
    - print its length and type using length() and typeof() functions


```{r}
   # Create the vector
   student_ages <- c(18, 21, 20, 22, 19)
   
   # Print the length and type
   length(student_ages)  # Output: 5
   typeof(student_ages)  # Output: "double"
```


2. **Create a factor** `student_grades` representing the grades of the same 5 students: "A", "B", "A", "C", "B".
    - print the factor levels, why do you use factors?


```{r}
   # Create the factor
   student_grades <- factor(c("A", "B", "A", "C", "B"))
   
   # Print the factor levels
   levels(student_grades)  # Output: "A" "B" "C"
   
```

Factors are useful here because grades are categorical data with a limited number of levels.


3. **Create a list** `student_info` that combines `student_ages`, `student_grades`, and another vector `student_names`.
    - access individual elements using both numeric and named indexing.


```{r}
   # Create the vector of student names
   student_names <- c("Aurelien", "Andrea", "Federica", "Bob", "Eva")
   
   # Create the list
   student_info <- list(names = student_names, ages = student_ages, grades = student_grades)
   
   # Access elements in the list
   student_info$names  
   student_info[[1]]
```

4. What is the difference between a vector and a factor? Why do you use a list instead of a vector when dealing with different types of data?

   - **Difference between vector and factor**: A vector is a basic data structure for storing homogeneous data types (e.g., numbers, characters), whereas a factor is specifically designed for categorical data.
   - **Why use a list instead of a vector?**: A list can store elements of different types (e.g., vectors, factors), making it more flexible than a vector.

#### Matrices and Arrays

1. **Create a matrix** `scores_matrix` representing the scores of 3 students in 3 subjects.
    - give row and column names using rownames() and colnames().
    - print the matrix and the score of a student that you prefer.
  

```{r}
   # Create the matrix
   scores_matrix <- matrix(c(75, 82, 78, 88, 90, 92, 60, 65, 70), nrow = 3, byrow = TRUE)
   
   # Set row and column names
   rownames(scores_matrix) <- c("Student 1", "Student 2", "Student 3")
   colnames(scores_matrix) <- c("Math", "English", "Science")
   
   # Print the matrix
   print(scores_matrix)
   
   # Access the score of Student 2 in Science
   scores_matrix["Student 2", "Science"]  
```

2. **Create an array** `student_array` to store the same data as in `scores_matrix` but for two semesters.
    - this 3-dimensional array, access "Student 3" scores in the first subject listed across both semesters.

```{r}
   # Create the array
   student_array <- array(c(
     75, 82, 78, 88, 90, 92, 60, 65, 70,  # Semester 1
     80, 84, 82, 91, 94, 93, 63, 68, 72   # Semester 2
   ), dim = c(3, 3, 2), 
   dimnames = list(
     c("Student 1", "Student 2", "Student 3"),
     c("Math", "English", "Science"),
     c("Semester 1", "Semester 2")
   ))
   
   # Access "Student 3's" scores in "Math" across both semesters
   student_array["Student 3", "Math", ]  # Output: 60 63
   
```

3. What are the differences between a matrix and an array? When would you use an array instead of a matrix?
    - A matrix is a 2-dimensional structure, while an array can have more than two dimensions.
    - Use an array when you need to store multi-dimensional data.

#### Data Frames, tibble and data.table

1. **Create a data frame** `students_df` combining all the data for 5 students: student_names, student_ages, student_grades, and another vector attendance containing their class attendance (in percentage): 85, 92, 78, 88, 90.
   - access the columns of attendance.
   - access the data for student 3.

```{r}
   # Create a data frame
   students_df <- data.frame(
     names = student_names,
     ages = student_ages,
     grades = student_grades,
     attendance = c(85, 92, 78, 88, 90)
   )
   
   # Access a specific column (student_names)
   students_df$names  

   # Access the data for "Andrea"
   students_df[students_df$names == "Andrea", ]  # Output: Data frame row for David
   
```

**Difference between a list and a data frame**: A list can contain elements of different types and structures, whereas a data frame is specifically a table where each column can have different types, but each column is a vector.
   
2. Why is a data frame more suitable for this kind of structured data than a matrix?
   -  Data frames are more suitable for structured data because they can contain heterogeneous data types and offer better data manipulation functions.


3. Convert the existing students_df data frame into a tibble using the as_tibble() function.
   - name the tibble students_tibble and print the data frame and the tibble.

```{r}
# Load the necessary package
library(tibble)

# Convert to tibble
students_tibble <- as_tibble(students_df)

# Print the tibble and data frame to observe differences
print(students_df)       # Standard data frame print
print(students_tibble)   # More concise print for tibbles

# Access a non-existent column in data frame and tibble
students_df$nonexistent  # Output: NULL
students_tibble$nonexistent  # Output: NULL with a warning
```

Tibbles provide a more readable print format and do not change types or convert strings to factors by default. Trying to access a non-existent column in a tibble gives a clear warning, unlike in data frames.

```{r}
# Load the necessary package
library(data.table)

# Convert to data.table
students_dt <- as.data.table(students_df)

# Selecting a column
print(students_dt[, ages])  # Output: Vector of ages

# Filtering rows where grades are "A"
print(students_dt[grades == "A"]) # Output: Rows with grade "A"

# Adding a new column with the average attendance
students_dt[, average_attendance := mean(attendance)]
print(students_dt)
```

data.table has a more concise syntax for data manipulation tasks, such as filtering and adding new columns, and it is significantly faster, especially with large datasets


## Additional Exercises for training

5. Use the output from the previous exercise. Compare the syntax of common data manipulation tasks (e.g., filtering, grouping, summarizing) between data.frame, tibble, and data.table.
   - Filter students with attendance greater than 85
   - Calculate mean age of students who passed
   
```{r}
library(dplyr)
# Filter students with attendance greater than 85
# Using data.frame
df_filtered <- students_df[students_df$attendance > 85, ]

# Using tibble
tb_filtered <- students_tibble %>% filter(attendance > 85)

# Using data.table
dt_filtered <- students_dt[attendance > 85]

# Calculate mean age of students who passed
# Using data.frame
mean(students_df$ages[students_df$grades <= "C"])

# Using tibble
students_tibble %>% filter(grades <= "C") %>% summarise(mean_age = mean(ages))

# Using data.table
students_dt[grades <= "C", mean(ages)]

# Add a new column 'grade_numeric'
# Using data.frame
students_df$grade_numeric <- as.numeric(
  factor(students_df$grades, levels = c("A", "B", "C"), labels = c(4, 3, 2)))

# Using tibble
students_tibble <- students_tibble %>% mutate(
  grade_numeric = as.numeric(factor(grades, levels = c("A", "B", "C"), labels = c(4, 3, 2))))

# Using data.table
students_dt[, grade_numeric := as.numeric(
  factor(grades, levels = c("A", "B", "C"), labels = c(4, 3, 2)))]

```


## Exercise E: Standard Deviation Function

1. During the lecture on Programming with Data we have step-by-step implemented our own R function to compute the mean. Now implement an R function called `my_sd` that computes the standard deviation, defined as: $SD = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}$. Hint: revisit the math operators section in the notes on the first exercise session.

```{r}
my_sd <- function(x) {
  n <- length(x)
  
  # First, check if n is greater than 1
  if(n <= 1) {
    warning("At least two numbers are required to compute standard deviation.")
    return(NA) # NA is returned if computation is not possible
  }
  
  mean_x <- mean(x)
  sum_squared_diff <- sum((x - mean_x)^2)
  sd <- sqrt(sum_squared_diff / (n - 1))
  
  return(sd)
}

```


2. Test your implementation by comparing its output with the already implemented `sd()` function.

```{r}
# Create a test vector
set.seed(2023) # setting seed to reproduce the same results (note that results may still  
               # vary across different devices and R-versions)
test_vector <- rnorm(10000)  # a random vector of 10000 numbers with mean zero and sd of 1

# Compare the outputs of the custom function and the built-in function
custom_sd <- my_sd(test_vector)
builtin_sd <- sd(test_vector)

print(paste("Custom standard deviation:", custom_sd))
print(paste("Built-in standard deviation:", builtin_sd))

```



## Exercise F: Standard Error Function

Building on the function written in Exercise D, implement a function to compute (the estimate of) the sample mean standard error as defined here: $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$. Hint: the idea is to use your own implementation (in Exercise D) to compute $SD$ as a first step of your standard error function.

Exercises D and E should illustrate how you can approach more complex problems by breaking them up in smaller problems and addressing each of these step-by-step. That is, when looking at the problem of computing $\text{SE}_{\bar{x}} = \frac{SD}{\sqrt{n}}$, you see that computing $SD$ is a part of the overall problem, which can be addressed separately.


```{r}
my_se <- function(x) {
  # Call the previously defined my_sd function to compute the standard deviation
  sd <- my_sd(x)
  
  n <- length(x)
  if(n <= 1) {
    warning("At least two numbers are required to compute standard error.")
    return(NA) # NA is returned if computation is not possible
  }
  
  se <- sd / sqrt(n)
  return(se)
}


# Test the function (base R has not built in function for this - we will test the 
# final output below)
test_se <- my_se(test_vector)
print(paste("Standard error computed by my function:", test_se))

```



## Exercise G: T-test

Implement a function for the one-sample [t-test](https://en.wikipedia.org/wiki/Student%27s_t-test):
$t = \frac{\bar{x} - \mu_0}{SE}$.

```{r}
my_t_test <- function(x, mu_0) {
  # Calculate the sample mean
  sample_mean <- mean(x)
  
  # Compute standard error using the function from Exercise E
  se <- my_se(x)
  
  n <- length(x)
  if(n <= 1) {
    warning("At least two numbers are required to compute the t-statistic.")
    return(NA) # NA is returned if computation is not possible
  }
  
  # Calculate the t-statistic
  t_statistic <- (sample_mean - mu_0) / se
  
  return(t_statistic)
}

# Testing against R

# Test the function with a sample vector and a test mean (mu_0)
mu_0_test <- 0  # This is the hypothesized mean (and the true mean of our data)

# Because our hypothesized value is correct, we should not reject, i.e.
# get a low t-statistic, well below the 5% significance threshold of +-1.96
t_test_result <- my_t_test(test_vector, mu_0_test)
print(paste("T-test result:", t_test_result))

# Now, use the built-in t-test function for comparison
builtin_t_test_result <- t.test(test_vector, mu = mu_0_test)$statistic

print(paste("Built-in T-test result:", builtin_t_test_result))
```

