---
title: 'Data Handling: Import, Cleaning and Visualisation'
subtitle: 'Exercise 4:<br>Data Import'
author: "Prof. Dr. Ulrich Matter"
date: "25/11/2022"
output:   
  ioslides_presentation:
    css: ../../style/ioslides.css
    template: ../../style/nologo_template.html
    mathjax: "default"
---



```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE}
options(width = 100)
library(knitr)
```


<style>
pre {
  font-size: 21px;
}
</style>

```{r echo = FALSE, warning=FALSE}
library(readxl)
library(knitr)
library(magrittr)
library(kableExtra)
```

# Recap

## Sources/formats in economics

- CSV (typical for rectangular/table-like data)
- Variants of CSV (tab-delimited, fix length etc.)
- XML and JSON (useful for complex/high-dimensional data sets)
- HTML (a markup language to define the structure and layout of webpages)
- Unstructured text

## Sources/formats in economics


- Excel spreadsheets (`.xls`)
- Formats specific to statistical software packages (SPSS: `.sav`, STATA: `.dat`, etc.)
- Built-in R datasets
- Binary formats


# Loading/Importing Rectangular Data

## Loading built-in datasets

In order to load such datasets, simply use the `data()`-function.


## Comma Separated Values (CSV)
- Use `read.csv()` (basic R distribution), which returns a `data.frame`
- Alternative: `read_csv()` (`readr`/`tidyr`-package), which returns a `tibble`. 
    - `readr` is fast and flexible!
    - Parsers in `read_csv()` guess the data *types*.

## Spreadsheets/Excel

R-package: `readxl` and function `read_excel()` to import data from an excel-sheet. 



## Data from other data analysis software 

- STATA, SPSS, etc.
- Additional packages needed:
    - `foreign`
    - `haven`
- Parsers (functions) for many foreign formats.
    - For example, `read_spss()` for SPSS' `.sav`-format.


## Importing Web Data Formats

- XML in R (tree structure!) with `library(xml2)` and `xml_doc`
    - Navigate sidewards and upwards
- JSON in R with `library(jsonlite)` and function `fromJSON()`.
    - Nested lists! 

## Explore data
To inspect the data after loading, use the `str()` and the `head()` functions!
